---
title: "Homework 1"
format: docx
execute: 
  echo: true
---

Please copy and paste your code and answers into this document and submit on Canvas.

1. Install R and RStudio as well as the packages described in Section 1.4 of of R for Data Science. Paste the code you used to install those packages below.

```{r}
#| eval: false
install.packages("tidyverse")
install.packages(c("arrow", "babynames", "curl", "duckdb", "gapminder", "ggrepel", "ggridges", "ggthemes", "hexbin", "janitor", "Lahman", "leaflet", "maps", "nycflights13", "openxlsx", "palmerpenguins", "repurrrsive", "tidymodels", "writexl"))
```

2.	Create a vector named `temp` with the average low temperatures in December for London, Shanghai, Sydney, Boston, Mexico City and Johannesburg which are 40, 38, 64, 28, 42 and 57° Fahrenheit. 

```{r}
temp <- c(40, 38, 64, 28, 42, 57)
```

3.	Create a second vector with the city names called `city`.

```{r}
city <- c("London", "Shanghai", "Sydney", "Boston", "Mexico City", "Johannesburg")
```

4.	Use the `setNames()` function to assign the city names to the corresponding temperatures using the two objects create above. (Hint: `help(setNames)`). You should overwrite your original `temp` vector with the named vector. Print the `temp` vector.

```{r}
temp <- setNames(object = temp, city)
temp
```

5.	Recall that you can use the `[]` operator directly following a vector name to select specific values contained within that vector (e.g. `temp[1]` should return `40`). Use this `[]` operator and the `sum()` function to calculate the sum of the temperatures for Sydney, Mexico City and Johannesburg.

```{r}
sum(temp[3], temp[5], temp[6])
```

6.	Create a vector of all positive even numbers smaller than 75 using the `seq()` function. (Hint: `help(seq)`)

```{r}
pos_even <- seq(2,75,2)
pos_even
```

7.	Create a vector of numbers. Let the first value of the vector be 4, with a maximum value < 80, adding numbers in increments of 2/3. How many elements are in this vector?

```{r}
num_1 <- seq(4,79.9,2/3)
length(num_1)
```

g.  Using your answer to the previous question, find another solution to create the same vector as the previous question.

```{r}
num_2 <- seq(4,79+1/3,length.out=114)
#was not really sure, if you had a specific way of solving this in your mind; I can think of many different solutions (e.g. loops etc.) but was not sure what we would expect from the course participants at this level
## if they do submit a more complex way like a loop to get the same result, I wanted to double check that we would give them full credit for this question?
```
